---
title: "Forest Cover Type Classification Project"
author: "Thomas Kolasa and Aravind Kolumum Raja"
output: html_document 
---

### The Dataset

We found the Forest Covertype in the UCI Machine Learning Repository that takes forestry data from the [Roosevel National Forest](https://tourbuilder.withgoogle.com/tour/ahJzfmd3ZWItdG91cmJ1aWxkZXJyEQsSBFRvdXIYgICAk5uU5AoM) in northern Colorado. The observations are taken from 30m by 30m patches of forest that are classified as one of seven forest types:

1. Spruce/Fir
2. Lodgepole Pine
3. Ponderosa Pine
4. Cottonwood/Willow
5. Aspen
6. Douglas-fir
7. Krummholz

Kaggle hosted the dataset in a competition with a training set of 15,120 observations and a test set of 565,892 observations. The relative size of the training and test sets makes classification of cover type a challenging problem.

```{r, echo=FALSE}
#setwd('C://Users/Aravind/Documents/GitHub/bootcamp004_project/Project4-Machinelearning/Aravind_thomas')
setwd('/Users/tkolasa/dev/nycdatascience/projects/bootcamp004_project/Project4-Machinelearning/Aravind_thomas')

library(ggplot2)
library(GGally)

forestdata = read.csv('train.csv', header=TRUE)

# adding a new variable for covernames based on cover type

forestdata$covername = 'a'
forestdata$covername[forestdata$Cover_Type==1] = 'Spruce-fir'
forestdata$covername[forestdata$Cover_Type==2] = 'Lodgepole Pine'
forestdata$covername[forestdata$Cover_Type==3] = 'Ponderosa Pine'
forestdata$covername[forestdata$Cover_Type==4] = 'Cottonwood-Willow'
forestdata$covername[forestdata$Cover_Type==5] = 'Aspen'
forestdata$covername[forestdata$Cover_Type==6] = 'Douglas-fir'
forestdata$covername[forestdata$Cover_Type==7] = 'Krummholz'

# Create single Wilderness_Area column
forestdata$Wilderness_Area = 0
for (i in 12:15) {
  forestdata$Wilderness_Area[forestdata[,i] == 1] = i-11  
}

#plotting the correlation matrix of the quantitative factors 
forestcorrelation=cor(forestdata[,2:11])
# 
forestdata$Hillshade_mean = (forestdata$Hillshade_9am + forestdata$Hillshade_3pm + forestdata$Hillshade_Noon) / 3
forestdata$Euclidean_Distance_To_Hydrology = (forestdata$Horizontal_Distance_To_Hydrology^2 + forestdata$Vertical_Distance_To_Hydrology^2)^.5

forestdata$Aspect[forestdata$Aspect == 360] = 0

rose_diagram_df = forestdata[c('Aspect', 'covername')]

rose_diagram_df['aspect_group'] = cut(rose_diagram_df$Aspect, breaks=c(-1,seq(20,360, by = 20)), labels=FALSE)

forestdata$aspect_group=rose_diagram_df$aspect_group
forestdata$aspect_group_shift= forestdata$aspect_group+3
forestdata$aspect_group_shift[forestdata$aspect_group_shift==19]= 1 
forestdata$aspect_group_shift[forestdata$aspect_group_shift==20]= 2 
forestdata$aspect_group_shift[forestdata$aspect_group_shift==21]= 3 

# Create single Soil_Type column
forestdata$Soil_Type = 0
for (i in 16:55) {
  forestdata$Soil_Type[forestdata[,i] == 1] = i-15  
}

################### converting to factors ########################

forestdata$Soil_Type=as.factor(forestdata$Soil_Type)
forestdata$Wilderness_Area=as.factor(forestdata$Wilderness_Area)

forestdata$covername=as.factor(forestdata$covername)
forestdata$Cover_Type=as.factor(forestdata$Cover_Type)
forestdata$aspect_group=as.factor(forestdata$aspect_group)
forestdata$aspect_group_shift=as.factor(forestdata$aspect_group_shift)

forestdata1= forestdata[,-c(22,30)]


for(i in 12:53)
{
  forestdata1[,i]= as.factor(forestdata1[,i])
  
}
forestdata1=forestdata1[,1:54]
```

```{r, echo=FALSE}

```