---
title: "Internet usage in the UK"
author: "Anna"
date: "23 Jan 2016"
output: ioslides_presentation
---


```{r, echo=FALSE, message=FALSE, warning=FALSE}
library(ggplot2)
library(dplyr)
library(xlsx)
library(RColorBrewer)
library(maps)
library(ggthemes)
library(data.table)
library(leaflet)
library(gridExtra)

age = read.xlsx("~/Downloads/internet-usage-london.xls", 3)
ethnicity = read.xlsx("~/Downloads/internet-usage-london.xls", 4)
pay = read.xlsx("~/Downloads/internet-usage-london.xls", 6)
characteristics = read.xlsx("~/Downloads/internet-usage-london.xls", 7)
geo = read.xlsx("~/Downloads/internet-usage-london.xls", 2)
geo = geo[!apply(geo, 1, FUN=function(x) any(is.na(x))),]
income = read.xlsx("~/Downloads/internet-usage-london.xls", 6)
jobs = read.xlsx("~/Downloads/internet-usage-london.xls", 5)
geo2015=filter(geo, Year == 2015)
geo2015=select(geo2015, Borough, epercent, ever.used.number)
gadm <- readRDS("/Users/bohun/Downloads/GBR_adm2.rds")
```
  

```{r, echo=FALSE, message=FALSE}
age$Year <- as.factor(age$Year)
temp = select(age, Age, epercent, npercent, ever.used.number, Year, Location)# %>%
#  filter(., Location=='London') 
tab = filter(temp, Age!="Working-age 16-64" & Age!="Total" )
totals = filter(temp, Age == "Total")  
```
  
  
##What factors affect internet usage?
- Age
- Ethnicity
- Income
- Occupation
- Location

Sample size is 333,000 people in the UK, 27,000 in London, over 33 boroughs from 2011-2015. 
If for a certain age range, internet usage depends categorically on income, a density map of users should give a rough indication of the higher income areas.  

Limitations: small time frame
  
##Total percentage of people using internet in the last five years
```{r, echo=FALSE, message=FALSE}
totals$Year = as.factor(totals$Year)
totalsplot=ggplot(data=totals, aes(x=as.numeric(totals$Year), y=epercent)) + geom_smooth(method=lm, aes(color=Location),se=FALSE)  +geom_point() + scale_x_continuous(breaks=1:nlevels(as.factor(totals$Year)),labels=c(2011:2015)) + xlab("Year") + ylab("Percent")
totalsplot + xlab("Year") + ylab("Percent") + ggtitle("Percent of the London and UK using internet since 2011")  
```


##How old are most internet users?
```{r, echo=FALSE, message=FALSE, fig.width=8, fig.height=5}
tab$Year <- factor(tab$Year, levels = 2015:2011)
age.use = ggplot(data=tab, aes(x=Age, y=npercent)) + geom_bar(stat='identity', aes(fill=Year), position = 'dodge')
plot1 = age.use+ scale_fill_brewer(palette = "Blues") + xlab("Age") + ylab("Percent") + theme_bw() + ggtitle("Percent of the UK which has never used internet")  
plot1
```



## Can we plot usage as a function of age?
```{r, echo=FALSE, message=FALSE, fig.width=8, fig.height=5}
midfunc = function(x)
  { mean(as.numeric(ifelse( x=="75+", 75, strsplit(as.character(x), split = "-"))[[1]]) )
  }
age.mids = mutate(tab, age.midpoint = sapply(tab$Age, midfunc )  ) 

age.slope = ggplot(data=age.mids, aes(x=age.midpoint, y=ever.used.number/10^6)) + geom_smooth(aes(color=tab$Year),se=FALSE)  

age.slope = age.slope + xlab("Age") + ylab("Number of users (millions)")  + scale_color_discrete(name = "Year") 
 
p.slope = ggplot(data=age.mids, aes(x=age.midpoint, y=epercent)) + geom_smooth(aes(color=tab$Year),se=FALSE)  
p.slope = p.slope + xlab("Age") + ylab("Percent of users") + theme(legend.position="none")

grid.arrange(age.slope, p.slope, ncol = 2)
```


##Changes over time
```{r, echo=FALSE, message=FALSE}
age$Age = as.factor(age$Age) 
totaluk = group_by(age, Age, Year) %>%
  summarise(., total = sum(ever.used.number), totalp = mean(epercent)) %>%
  filter(., Age!="Working-age 16-64" & Age!="Total") 
#Drop levels so they do not appear in the later analysis.
totaluk$Age = factor(totaluk$Age)
totaluk$totalscaled = totaluk$total/10^6
totaluk$Year = as.numeric(as.character(totaluk$Year))
plotuk = ggplot(totaluk, aes(x=as.numeric(Year), y=totaluk$totalscaled) ) + geom_smooth(method=lm, aes(color=totaluk$Age), se=FALSE)  +geom_point()
plotuk + xlab("Year") + ylab("Number of users (millions)") + ggtitle("Comparisons of increase by age")  +  scale_color_discrete( name = "Age group") 
```

## Which group is growing the fastest?
- Need to normalise
```{r}
dat2 = data.table(x=totaluk$Year, y=totaluk$totalp, group = totaluk$Age)
dat2[,list(intercept = coef(lm(y~x))[1], coef = coef(lm(y~x))[2]), by=group]
```



## Income
```{r, echo=FALSE, message=FALSE, fig.width=9, fig.height=5}
income.uk = group_by(income, gross.weekly.pay) %>%
  summarise(., ever.used = sum(ever.used.number), ever.p = mean(epercent))
x =income.uk$gross.weekly.pay
#Remove all the pound and special characters from income
x = gsub("[\\£]","", x)  
x = gsub("[\\<]","",x)
x = gsub("[\\+]","", x)
income.uk$temp = x
sal = function(x)
  { as.numeric(ifelse( x=="2000", 2000, ifelse(x=="200", 200, strsplit(as.character(x), split = "-")))[[1]]) 
  }
income.uk$sal = sapply(income.uk$temp, sal )
income.uk = mutate(income.uk, annual = sapply(income.uk$sal, function(x) {x*52}))
income.uk = mutate(income.uk, annualsal = sapply(income.uk$annual, function(x) {paste(x, collapse='-')})) 
income.uk = mutate(income.uk, annualmedian = sapply(income.uk$annual, function(x) mean(as.numeric(x)) )) 
income.uk[1,7] = "<10400"
income.uk[5,7] = "104000+"

income.per = ggplot(data=income.uk, aes(x=reorder(annualsal, annualmedian), y=ever.p)) + geom_bar(stat='identity', aes(fill=as.factor(annualsal))) 
income.per + xlab("Income range (GBP)") + ylab("Percent of users") + ggtitle("Internet users by income range (percentage) in the UK")  + scale_fill_discrete( name = "Income Range") + coord_cartesian(ylim = c(90, 100)) 
```

## But in absolute value...
```{r, echo=FALSE, message=FALSE, fig.width=9, fig.height=5}
income.plot = ggplot(data=income.uk, aes(x=reorder(annualsal, annualmedian), y=ever.used/10^6)) + geom_bar(stat='identity', aes(fill=as.factor(annualsal))) 
income.plot + xlab("Income range (GBP)") + ylab("Number of users (millions)") + ggtitle("Internet users by income range in the UK") + scale_fill_discrete( name = "Income Range") 
```


##Population map
- We want to visualize number of users on map, but the dataset is too small and missing areas in the UK
- Download UK data and focus on London

##Map by numbers
```{r, echo=FALSE, message=FALSE}

geo2015=filter(geo, Year == 2015)
geo2015=select(geo2015, Borough, epercent, ever.used.number)
gadm <- readRDS("/Users/bohun/Downloads/GBR_adm2.rds")

gadm$enumber =  ifelse(gadm$NAME_2 %in% geo2015$Borough, geo2015$ever.used.number, NA)
london <- gadm[gadm$NAME_2 %in% geo2015$Borough,]

london$colornum2 =((london$enumber -  min(london$enumber))/(max(london$enumber)-min(london$enumber))*33)


polygon_popup2 = paste0("<strong>", london$NAME_2, "</strong>", "</br>", london$enumber)

pal <- colorQuantile("Greens", NULL, n = 10)

leaflet(london) %>%
  setView(lat = 51.508, lng = -0.129158, zoom = 10) %>%
  addProviderTiles("CartoDB.Positron") %>%
   addPolygons( data = london, fillColor = ~pal(colornum2), 
   fillOpacity = 0.6,
   weight = 2,
   color = "white",
   popup = polygon_popup2
  ) %>%
    addLegend("bottomright", pal = colorNumeric( palette = pal, domain = london$enumber ), values = ~enumber,
    title = "Number of users",
    opacity = 1)
```


##Density map
```{r, results = 'asis', echo=FALSE, message= FALSE}   
gadm$epercent = ifelse(gadm$NAME_2 %in% geo2015$Borough, geo2015$epercent, NA)

london <- gadm[gadm$NAME_2 %in% geo2015$Borough,]
#create a subdata frame from gadm featuring only the areas for which we have data

london$colornum = round((london$epercent -  min(london$epercent))/(max(london$epercent)-min(london$epercent))*33)
#create the palette

polygon_popup = paste0("<strong>", london$NAME_2, "</strong>", "</br>", london$epercent)
#create a legend with labels showing the percent

pal <- colorQuantile("Blues", NULL, n = 10)
#will be 'broken up' into shades depending on colornum

leaflet(london) %>%
  setView(lat = 51.508, lng = -0.129158, zoom = 10) %>%
  addProviderTiles("CartoDB.Positron") %>%
   addPolygons( data = london, fillColor = ~pal(colornum), 
   fillOpacity = 0.6,
   weight = 2,
   color = "white",
   popup = polygon_popup
  ) %>%
    addLegend("bottomright", pal = colorNumeric( palette = pal, domain = london$epercent ), values = ~epercent,
    title = "Percentage of users",
    opacity = 1)
 

```


